---
title: "The Effect of Furlough on Cognitive Mental Health: An IV Approach"
output: html_document
date: "2025-02-19"
---

# ============================================================
# SECTION 0: SETUP — Libraries & Environment
# ============================================================

```{r setup, message=FALSE}
library(tidyverse)
library(haven)
library(broom)
library(tidyr)
library(stringr)
library(purrr)
library(lmtest)
library(sandwich)
library(AER)
library(fixest)
library(psych)
library(tableone)
library(modelsummary)

rm(list = ls())
```


# ============================================================
# SECTION 1: DATA LOADING
# ============================================================

```{r load-data, message=FALSE}
# --- UKHLS COVID-19 survey waves (April–September 2020) ---
april <- read_sav("~/Downloads/UKDA-8644 covid-spss/spss/spss25/ca_indresp_w.sav")
may   <- read_sav("~/Downloads/UKDA-8644 covid-spss/spss/spss25/cb_indresp_w.sav")
jun   <- read_sav("~/Downloads/UKDA-8644 covid-spss/spss/spss25/cc_indresp_w.sav")
jul   <- read_sav("~/Downloads/UKDA-8644 covid-spss/spss/spss25/cd_indresp_w.sav")
sep   <- read_sav("~/Downloads/UKDA-8644 covid-spss/spss/spss25/ce_indresp_w.sav")

covid_list <- list(ca = april, cb = may, cc = jun, cd = jul, ce = sep)
```


# ============================================================
# SECTION 2: WAVE CLEANING & HARMONISATION
# ============================================================

```{r clean-waves}
# --- Variables to retain after harmonisation ---
harmonized_vars <- c(
  "pidp", "age", "sex", "couple", "sempderived", "blwork",
  "netpay_amount", "netpay_period", "keyworksector",
  "treated", "employed", "wave", "post",
  "scghqa", "scghqb", "scghqc", "scghqd", "scghqe", "scghqf",
  "scghqg", "scghqh", "scghqi", "scghqj", "scghqk", "scghql",
  "furlough", "newfurlough", "stillfurl",
  "blwah"  # instrument: ability to work from home pre-pandemic
)

# --- Function: strip wave prefix, harmonise sex/keyworksector, add wave label ---
clean_wave <- function(df, prefix, wave_label) {
  df <- df %>%
    rename_with(~str_remove(., paste0("^", prefix, "_")))

  # Harmonise sex if only sex_cv is present
  if (!"sex" %in% names(df) && "sex_cv" %in% names(df)) {
    df$sex <- df$sex_cv
  }

  df <- df %>%
    mutate(
      treated  = if ("keyworksector" %in% names(df)) ifelse(keyworksector == 9 | is.na(keyworksector), 1, 0) else NA,
      employed = ifelse(sempderived %in% c(1,2,3), 1, ifelse(sempderived %in% c(0,4), NA, 0)),
      sex      = case_when(sex == 1 ~ 1, sex == 2 ~ 0, TRUE ~ NA_real_),
      wave     = wave_label,
      post     = 1
    ) %>%
    select(-any_of(c("sex_cv", "keyworker")))

  df %>% select(any_of(harmonized_vars))
}

# Apply to each wave
april <- clean_wave(covid_list$ca, "ca", "april")
may   <- clean_wave(covid_list$cb, "cb", "may")
jun   <- clean_wave(covid_list$cc, "cc", "june")
jul   <- clean_wave(covid_list$cd, "cd", "july")
sep   <- clean_wave(covid_list$ce, "ce", "sep")

# Combine into single panel
covid_combined <- bind_rows(
  list(april = april, may = may, jun = jun, jul = jul, sep = sep),
  .id = "wave_short"
)
```


# ============================================================
# SECTION 3: VARIABLE RECODING
# ============================================================

```{r recode-variables}
# --- Clean numeric variables: set negatives & sentinel codes to NA ---
covid_combined <- covid_combined %>%
  mutate(
    netpay_amount = ifelse(netpay_amount %in% c(-9,-8,-2,-1), NA, netpay_amount),
    age           = ifelse(age           %in% c(-9,-8,-2,-1), NA, age)
  ) %>%
  mutate(across(everything(), ~ifelse(. < 0, NA, .)))

# --- Furlough status: harmonise across wave-specific variable names ---
# furlough / newfurlough: 1 = furloughed, 2 = not furloughed
# stillfurl: 1–2 = still furloughed, 3–5 = returned/left
add_furlough_status <- function(df) {
  furlough_status <- rep(NA_real_, nrow(df))

  if ("furlough" %in% names(df)) {
    furlough_status[df$furlough == 1] <- 1
    furlough_status[df$furlough == 2] <- 0
  }
  if ("newfurlough" %in% names(df)) {
    furlough_status[df$newfurlough == 1] <- 1
    furlough_status[df$newfurlough == 2] <- 0
  }
  if ("stillfurl" %in% names(df)) {
    furlough_status[df$stillfurl %in% c(1,2)] <- 1
    furlough_status[df$stillfurl %in% c(3,4,5)] <- 0
  }

  df$furlough_status <- furlough_status
  return(df)
}

covid_combined <- add_furlough_status(covid_combined)

# --- Instrument: inability to work from home (blwah == 4) ---
covid_combined <- covid_combined %>%
  mutate(low_wfh = ifelse(blwah == 4, 1, 0))

# --- Year indicator ---
covid_combined <- covid_combined %>%
  mutate(year = ifelse(wave_short %in% c("ca","cb","cc","cd","ce"), 2020, NA_integer_))

# --- Income: clean and log-transform net pay ---
covid_combined <- covid_combined %>%
  mutate(
    netpay_clean = ifelse(netpay_amount > 0, netpay_amount, NA),
    log_netpay   = log(netpay_clean),
    male         = ifelse(sex == 1, 1, 0)
  )

# --- Age groups ---
covid_combined <- covid_combined %>%
  mutate(age_group = case_when(
    age < 35             ~ "young",
    age >= 35 & age <= 50 ~ "middle",
    age > 50             ~ "older"
  ))

# --- Income tertiles ---
covid_combined <- covid_combined %>%
  mutate(
    income_group = ntile(log_netpay, 3),
    low_income   = if_else(income_group == 1, 1, 0),
    mid_income   = if_else(income_group == 2, 1, 0),
    high_income  = if_else(income_group == 3, 1, 0)
  )
```


# ============================================================
# SECTION 4: MENTAL HEALTH OUTCOMES — GHQ SCORING
# ============================================================

```{r ghq-scoring}
ghq_vars <- c("scghqa","scghqb","scghqc","scghqd","scghqe","scghqf",
              "scghqg","scghqh","scghqi","scghqj","scghqk","scghql")

# --- Binary (Caseness) scoring: 0-0-1-1 ---
# Responses 1–2 → 0 (no problem), 3–4 → 1 (problem present)
covid_combined <- covid_combined %>%
  mutate(across(all_of(ghq_vars),
    ~case_when(.x %in% c(1,2) ~ 0, .x %in% c(3,4) ~ 1, TRUE ~ NA_real_),
    .names = "bin_{.col}"))

# --- Total GHQ-12 score (0–12) ---
covid_combined <- covid_combined %>%
  rowwise() %>%
  mutate(ghqscore = sum(c_across(starts_with("bin_")), na.rm = FALSE)) %>%
  ungroup()

# --- Domain sub-scores ---
covid_combined <- covid_combined %>%
  mutate(
    ghq_cognitive     = rowSums(select(., bin_scghqa, bin_scghqd), na.rm = TRUE),
    ghq_stress        = rowSums(select(., bin_scghqb, bin_scghqe, bin_scghqf), na.rm = TRUE),
    ghq_functionality = rowSums(select(., bin_scghqc, bin_scghqg), na.rm = TRUE),
    ghq_depression    = rowSums(select(., bin_scghqh, bin_scghqj, bin_scghqk), na.rm = TRUE),
    ghq_wellbeing     = bin_scghql
  )
```


# ============================================================
# SECTION 5: BASELINE OLS
# ============================================================

```{r ols-baseline}
# --- OLS: furlough on cognitive mental health (naive, no IV) ---
ols_model <- lm(
  ghq_cognitive ~ furlough_status + age + sex + couple + log_netpay + treated + factor(wave_short),
  data = covid_combined
)

# Heteroscedasticity-robust standard errors
coeftest(ols_model, vcov = vcovHC(ols_model, type = "HC1"))
```


# ============================================================
# SECTION 6: INSTRUMENTAL VARIABLES STRATEGY
# ============================================================

```{r first-stage}
# --- First stage: low_wfh → furlough_status ---
# Rule of thumb: F-statistic > 10 indicates a strong instrument
first_stage <- lm(
  furlough_status ~ low_wfh + age + sex + couple + log_netpay + treated + factor(wave_short),
  data = covid_combined
)

coeftest(first_stage, vcov = vcovHC(first_stage, type = "HC1"))
summary(first_stage)$fstatistic  # Check instrument strength
```

```{r exclusion-restriction-check}
# --- Reduced form: instrument should only affect outcome through furlough ---
# If low_wfh is significant here after controlling for all covariates,
# there may be a direct pathway violating the exclusion restriction
exclusion_check <- lm(
  ghq_cognitive ~ low_wfh + age + sex + couple + log_netpay + treated + factor(wave_short),
  data = covid_combined
)
coeftest(exclusion_check, vcov = vcovHC(exclusion_check, type = "HC1"))
```

```{r main-iv-model}
# --- Main 2SLS model: effect of furlough on cognitive mental health ---
iv_model <- ivreg(
  ghq_cognitive ~ furlough_status + age + sex + couple + log_netpay + treated + factor(wave_short) |
    low_wfh + age + sex + couple + log_netpay + treated + factor(wave_short),
  data = covid_combined
)

summary(iv_model, diagnostics = TRUE)
```


# ============================================================
# SECTION 7: BALANCE CHECK
# ============================================================

```{r balance-check}
# --- Assess covariate balance across instrument groups (low_wfh = 0 vs 1) ---
# Good instrument should be unrelated to pre-treatment characteristics
balance_vars  <- c("age", "male", "couple", "employed", "log_netpay", "treated")
balance_table <- CreateTableOne(vars = balance_vars, strata = "low_wfh",
                                data = covid_combined, test = TRUE)
print(balance_table, smd = TRUE)
```


# ============================================================
# SECTION 8: ITEM-LEVEL IV — EFFECTS ON EACH GHQ DIMENSION
# ============================================================

```{r ghq-item-level-iv}
# GHQ item labels for readable output
label_lookup <- tibble(
  outcome = c("bin_scghqa","bin_scghqb","bin_scghqc","bin_scghqd","bin_scghqe","bin_scghqf",
              "bin_scghqg","bin_scghqh","bin_scghqi","bin_scghqj","bin_scghqk","bin_scghql"),
  label   = c("Concentration","Loss of sleep","Playing useful role","Decision-making",
              "Under strain","Overcoming difficulties","Enjoy daily activities","Face problems",
              "Unhappy/depressed","Losing confidence","Worthlessness","General happiness"),
  domain  = c("Cognitive","Stress","Functionality","Cognitive","Stress","Stress",
              "Functionality","Functionality","Depression","Depression","Depression","Wellbeing")
)

# Run IV for each GHQ item and extract furlough coefficient
ghq_iv_results <- map_dfr(label_lookup$outcome, function(var) {
  formula_iv <- as.formula(paste0(
    var, " ~ furlough_status + age + sex + couple | low_wfh + age + sex + couple"
  ))
  ivreg(formula_iv, data = covid_combined) %>%
    tidy() %>%
    filter(term == "furlough_status") %>%
    mutate(outcome = var, sig = p.value < 0.05)
}) %>%
  left_join(label_lookup, by = "outcome")

# --- Plot: IV estimates by GHQ domain ---
ggplot(ghq_iv_results, aes(x = estimate, y = fct_rev(label), color = domain)) +
  geom_point(size = 2) +
  geom_errorbarh(aes(xmin = estimate - 1.96 * std.error,
                     xmax = estimate + 1.96 * std.error)) +
  geom_vline(xintercept = 0, linetype = "dashed", alpha = 0.5) +
  facet_wrap(~domain, scales = "free_y", ncol = 1) +
  labs(title = "IV Estimates: Effect of Furlough on Mental Health by GHQ Domain",
       x = "Effect of Furlough", y = NULL, color = "GHQ Domain") +
  theme_minimal()
```


# ============================================================
# SECTION 9: HETEROGENEITY ANALYSIS
# ============================================================

```{r heterogeneity-by-gender}
# --- By gender ---
iv_male <- ivreg(
  ghq_cognitive ~ furlough_status + age + couple + factor(wave_short) |
    low_wfh + age + couple + factor(wave_short),
  data = covid_combined %>% filter(sex == 1)
)

iv_female <- ivreg(
  ghq_cognitive ~ furlough_status + age + couple + factor(wave_short) |
    low_wfh + age + couple + factor(wave_short),
  data = covid_combined %>% filter(sex == 0)
)

summary(iv_male)
summary(iv_female)
```

```{r heterogeneity-by-age}
# --- By age group ---
run_iv_age <- function(age_grp) {
  ivreg(
    ghq_cognitive ~ furlough_status + sex + couple + factor(wave_short) |
      low_wfh + sex + couple + factor(wave_short),
    data = covid_combined %>% filter(age_group == age_grp)
  )
}

iv_young  <- run_iv_age("young")
iv_middle <- run_iv_age("middle")
iv_older  <- run_iv_age("older")

summary(iv_young)
summary(iv_middle)
summary(iv_older)
```

```{r heterogeneity-by-couple-status}
# --- By relationship status ---
iv_couple <- ivreg(
  ghq_cognitive ~ furlough_status + age + sex + factor(wave_short) |
    low_wfh + age + sex + factor(wave_short),
  data = covid_combined %>% filter(couple == 1)
)

iv_single <- ivreg(
  ghq_cognitive ~ furlough_status + age + sex + factor(wave_short) |
    low_wfh + age + sex + factor(wave_short),
  data = covid_combined %>% filter(couple == 0)
)

summary(iv_couple)
summary(iv_single)
```

```{r heterogeneity-by-income}
# --- By income tertile (interacted within a single model) ---
# Base group: low income. Interactions capture differential effects for mid & high income.
covid_combined <- covid_combined %>%
  mutate(
    furlough_x_midinc  = furlough_status * mid_income,
    furlough_x_highinc = furlough_status * high_income,
    low_wfh_x_midinc   = low_wfh * mid_income,
    low_wfh_x_highinc  = low_wfh * high_income
  )

iv_income <- ivreg(
  ghq_cognitive ~ furlough_status + furlough_x_midinc + furlough_x_highinc +
    age + sex + couple + factor(wave_short) |
    low_wfh + low_wfh_x_midinc + low_wfh_x_highinc +
    age + sex + couple + factor(wave_short),
  data = covid_combined
)
summary(iv_income, diagnostics = TRUE)
```

```{r heterogeneity-multi-factor-interaction}
# --- Combined demographic interactions (age >40, male, in couple) ---
covid_combined <- covid_combined %>%
  mutate(
    over_40              = if_else(age > 40, 1, 0),
    in_couple            = if_else(couple == 1, 1, 0),
    furlough_x_over40   = furlough_status * over_40,
    furlough_x_male     = furlough_status * male,
    furlough_x_couple   = furlough_status * in_couple,
    low_wfh_x_over40    = low_wfh * over_40,
    low_wfh_x_male      = low_wfh * male,
    low_wfh_x_couple    = low_wfh * in_couple
  )

iv_hetero_multi <- ivreg(
  ghq_cognitive ~ furlough_status + furlough_x_over40 + furlough_x_male + furlough_x_couple +
    age + sex + couple + factor(wave_short) |
    low_wfh + low_wfh_x_over40 + low_wfh_x_male + low_wfh_x_couple +
    age + sex + couple + factor(wave_short),
  data = covid_combined
)
summary(iv_hetero_multi)
```


# ============================================================
# SECTION 10: ROBUSTNESS CHECKS
# ============================================================

```{r robustness-extra-controls}
# --- Add extra controls (netpay, treated) to main specification ---
iv_extra_controls <- ivreg(
  ghq_cognitive ~ furlough_status + age + sex + couple + log_netpay + treated + factor(wave_short) |
    low_wfh + age + sex + couple + log_netpay + treated + factor(wave_short),
  data = covid_combined
)
summary(iv_extra_controls)
```

```{r robustness-placebo-outcome}
# --- Placebo test: random fake outcome — furlough should have no effect ---
set.seed(123)
covid_combined$fake_outcome <- rnorm(nrow(covid_combined))

iv_placebo <- ivreg(
  fake_outcome ~ furlough_status + age + sex + couple + log_netpay + treated + factor(wave_short) |
    low_wfh + age + sex + couple + log_netpay + treated + factor(wave_short),
  data = covid_combined
)
summary(iv_placebo)
```

```{r robustness-working-age-only}
# --- Restrict to working age (18–65) ---
iv_working_age <- ivreg(
  ghq_cognitive ~ furlough_status + age + sex + couple + factor(wave_short) |
    low_wfh + age + sex + couple + factor(wave_short),
  data = covid_combined %>% filter(age >= 18 & age <= 65)
)
summary(iv_working_age)
```

```{r robustness-drop-wave}
# --- Drop May wave: check no single wave drives results ---
iv_no_may <- ivreg(
  ghq_cognitive ~ furlough_status + age + sex + couple + factor(wave_short) |
    low_wfh + age + sex + couple + factor(wave_short),
  data = covid_combined %>% filter(wave_short != "may")
)
summary(iv_no_may)
```

```{r robustness-alternative-outcomes}
# --- Alternative outcomes: depression and stress sub-indices ---
iv_depression <- ivreg(
  ghq_depression ~ furlough_status + age + sex + couple + log_netpay + treated + factor(wave_short) |
    low_wfh + age + sex + couple + log_netpay + treated + factor(wave_short),
  data = covid_combined
)
summary(iv_depression)

iv_stress <- ivreg(
  ghq_stress ~ furlough_status + age + sex + couple + log_netpay + treated + factor(wave_short) |
    low_wfh + age + sex + couple + log_netpay + treated + factor(wave_short),
  data = covid_combined
)
summary(iv_stress)
```
