---
title: "Conflict and Support for Democracy in Nigeria"
output: html_document
date: "2025-02-17"
---

# ============================================================
# SECTION 0: SETUP — Libraries & Environment
# ============================================================

```{r setup, message=FALSE}
library(tidyverse)
library(sf)
library(rnaturalearth)
library(rnaturalearthhires)
library(rnaturalearthdata)
library(haven)
library(tidyr)
library(dplyr)
library(fixest)
library(modelsummary)
library(knitr)

rm(list = ls())
```


# ============================================================
# SECTION 1: DATA LOADING
# ============================================================

```{r load-survey-data, message=FALSE}
# --- Afrobarometer survey rounds ---
df_1999 <- read_sav("~/Desktop/Dissertation /Nigeria Round 1 data (1999)")
df_2001 <- read_sav("~/Desktop/Dissertation /Nigeria Round 1.5 data (2001)")
df_2003 <- read_sav("~/Desktop/Dissertation /Nigeria Round 2 data (2003)")
df_2005 <- read_sav("~/Desktop/Dissertation /Nigeria Round 3 data (2005)")
df_2008 <- read_sav("~/Desktop/Dissertation /Nigeria Round 4 data (2008)")
df_2012 <- read_sav("~/Desktop/Dissertation /Nigeria Round 5 data (2013)")
df_2015 <- read_sav("~/Desktop/Dissertation /Nigeria Round 6 data (2015)")
df_2018 <- read_sav("~/Desktop/Dissertation /Nigeria Round 7 data (2017)")
df_2021 <- read_sav("~/Desktop/Dissertation /Nigeria Round 8 data (2021)")
df_2022 <- read_sav("~/Desktop/Dissertation /Nigeria Round 9 data (2022)")

# --- ACLED conflict data ---
crime <- read.csv("~/Desktop/Dissertation /Africa_1997-2025_Jan31.csv")
```


# ============================================================
# SECTION 2: VARIABLE MAPPING DICTIONARY
# Each entry maps standardised variable names to raw survey column names
# ============================================================

```{r column-dictionary}
column_mappings <- list(
  "1999" = c("ID" = "RESPNO",  "gender" = "GENDER",  "age" = "Q1",   "education" = "Q4",
             "employment" = "Q9",   "state" = "Q101",  "Urban/Rural" = "Q102",
             "support for democracy" = "Q36"),
  "2001" = c("ID" = "respno",  "gender" = "q96",    "age" = "q80",  "education" = "q84",
             "employment" = "q89",  "state" = "state", "Urban/Rural" = "urbrur",
             "support for democracy" = "supdemx1"),
  "2003" = c("ID" = "respno",  "gender" = "q96",    "age" = "q80",  "education" = "q84",
             "employment" = "q89",  "state" = "regon2","Urban/Rural" = "urbrur",
             "support for democracy" = "q38"),
  "2005" = c("ID" = "RESPNO",  "gender" = "Q101",   "age" = "Q1",   "education" = "Q90",
             "employment" = "Q94",  "state" = "REGION","Urban/Rural" = "URBRUR",
             "support for democracy" = "Q37"),
  "2008" = c("ID" = "RESPNO",  "gender" = "Q101",   "age" = "Q1",   "education" = "Q89",
             "employment" = "Q94",  "state" = "REGION","Urban/Rural" = "URBRUR",
             "support for democracy" = "Q30"),
  "2012" = c("ID" = "RESPNO",  "gender" = "Q101",   "age" = "Q113", "education" = "Q117",
             "employment" = "Q96",  "state" = "REGION","Urban/Rural" = "URBRUR",
             "support for democracy" = "Q32"),
  "2015" = c("ID" = "RESPNO",  "gender" = "Q101",   "age" = "Q1",   "education" = "Q97",
             "employment" = "Q95A", "state" = "REGION","Urban/Rural" = "URBRUR",
             "support for democracy" = "Q21"),
  "2018" = c("ID" = "RESPNO",  "gender" = "Q101",   "age" = "Q1",   "education" = "Q97",
             "employment" = "Q94",  "state" = "REGION","Urban/Rural" = "URBRUR",
             "support for democracy" = "Q28"),
  "2021" = c("ID" = "RESPNO",  "gender" = "Q101",   "age" = "Q1",   "education" = "Q97",
             "employment" = "Q95A", "state" = "REGION","Urban/Rural" = "URBRUR",
             "support for democracy" = "Q21"),
  "2022" = c("ID" = "RESPNO",  "gender" = "Q101",   "age" = "Q1",   "education" = "Q94",
             "employment" = "Q93A", "state" = "REGION","Urban/Rural" = "URBRUR",
             "support for democracy" = "Q23")
)
```


# ============================================================
# SECTION 3: SURVEY DATA CLEANING & STANDARDISATION
# ============================================================

```{r rename-and-tag-year}
# --- Function: rename columns using dictionary & add year tag ---
rename_correctly <- function(df, year) {
  mapping <- column_mappings[[as.character(year)]]
  if (!is.null(mapping)) {
    df <- df %>% rename(all_of(mapping))
  }
  df %>% mutate(year = year)
}

# Apply renaming to each wave
df_1999 <- rename_correctly(df_1999, 1999)
df_2001 <- rename_correctly(df_2001, 2001)
df_2003 <- rename_correctly(df_2003, 2003)
df_2005 <- rename_correctly(df_2005, 2005)
df_2008 <- rename_correctly(df_2008, 2008)
df_2012 <- rename_correctly(df_2012, 2012)
df_2015 <- rename_correctly(df_2015, 2015)
df_2018 <- rename_correctly(df_2018, 2018)
df_2021 <- rename_correctly(df_2021, 2021)
df_2022 <- rename_correctly(df_2022, 2022)
```

```{r select-relevant-columns}
# --- Function: keep only the 9 harmonised variables ---
select_relevant <- function(df) {
  df %>%
    select("ID", "support for democracy", "Urban/Rural",
           "age", "gender", "education", "employment", "state", "year")
}

df_1999 <- select_relevant(df_1999)
df_2001 <- select_relevant(df_2001)
df_2003 <- select_relevant(df_2003)
df_2005 <- select_relevant(df_2005)
df_2008 <- select_relevant(df_2008)
df_2012 <- select_relevant(df_2012)
df_2015 <- select_relevant(df_2015)
df_2018 <- select_relevant(df_2018)
df_2021 <- select_relevant(df_2021)
df_2022 <- select_relevant(df_2022)
```

```{r standardise-state-labels}
# --- Standardise state factor labels for each wave ---
# Waves 1999 & 2001: levels 1–38
state_labels_1_38 <- c(
  "Abia", "Adamawa", "Akwa-Ibom", "Anambra", "Bauchi", "Bayelsa", "Benue", "Borno",
  "Cross River", "Delta", "Ebonyi", "Edo", "Ekiti", "Enugu", "FCT", "Gombe", "Imo",
  "Jigawa", "Kaduna", "Kano", "Katsina", "Kebbi", "Kogi", "Kwara", "Lagos",
  "Nasarawa", "Niger", "Ogun", "Ondo", "Osun", "Oyo", "Plateau", "Rivers",
  "Sokoto", "Taraba", "Yobe", "Zamfara", "No Answer"
)

df_1999$state <- factor(df_1999$state, levels = 1:38, labels = state_labels_1_38)
df_2001$state <- factor(df_2001$state, levels = 1:38, labels = state_labels_1_38)

# Wave 2005: levels 340–359, 550–560
df_2005$state <- factor(df_2005$state,
  levels = c(340:359, 550:560),
  labels = c("Lagos", "Ogun", "Oyo", "Osun", "Ondo", "Ekiti", "Enugu", "Anambra", "Imo",
             "Abia", "Akwa-Ibom", "Bayelsa", "Cross-River", "Delta", "Edo", "Rivers", "Kano",
             "Sokoto", "Kaduna", "Katsina", "Zamfara", "Bauchi", "Borno", "Adamawa",
             "Taraba", "Plateau", "Benue", "Kogi", "Kwara", "Niger", "FCT"))

# Wave 2008: levels 620–645 (26 states only)
df_2008$state <- factor(df_2008$state,
  levels = 620:645,
  labels = c("Abia", "Adamawa", "Akwa Ibom", "Anambra", "Bauchi", "Bayelsa", "Benue", "Borno",
             "Cross River", "Delta", "Ebonyi", "Edo", "Ekiti", "Enugu", "FCT", "Gombe", "Imo",
             "Jigawa", "Kaduna", "Kano", "Katsina", "Kebbi", "Kogi", "Kwara", "Lagos", "Nasarawa"))

# Waves 2012–2022: levels 620–656 (all 37 states + FCT)
state_labels_620_656 <- c(
  "Abia", "Adamawa", "Akwa-Ibom", "Anambra", "Bauchi", "Bayelsa", "Benue", "Borno",
  "Cross-River", "Delta", "Ebonyi", "Edo", "Ekiti", "Enugu", "FCT", "Gombe", "Imo",
  "Jigawa", "Kaduna", "Kano", "Katsina", "Kebbi", "Kogi", "Kwara", "Lagos", "Nasarawa",
  "Niger", "Ogun", "Ondo", "Osun", "Oyo", "Plateau", "Rivers", "Sokoto", "Taraba",
  "Yobe", "Zamfara"
)

for (df_name in c("df_2012", "df_2015", "df_2018", "df_2021", "df_2022")) {
  df <- get(df_name)
  df$state <- factor(df$state, levels = 620:656, labels = state_labels_620_656)
  assign(df_name, df)
}
```

```{r standardise-democracy-variable}
# --- 1999 only: recode support for democracy so 1=low, 3=high (consistent with other waves) ---
df_1999 <- df_1999 %>%
  mutate(`support for democracy` = case_when(
    `support for democracy` == 1 ~ 3,
    `support for democracy` == 3 ~ 1,
    `support for democracy` == 2 ~ 2,
    TRUE ~ NA_real_
  ))
```

```{r combine-all-waves}
# --- Convert all columns to character before binding (avoids type conflicts) ---
convert_to_character <- function(df) {
  df %>% mutate(across(everything(), as.character))
}

df_list <- list(df_1999, df_2001, df_2003, df_2005, df_2008,
                df_2012, df_2015, df_2018, df_2021, df_2022)
df_list <- lapply(df_list, convert_to_character)

survey <- bind_rows(df_list)
```

```{r clean-combined-survey}
# --- Filter out missing, refused, and don't know responses ---
survey <- survey %>%
  filter(
    !`support for democracy` %in% c(NA, "8", "9"),
    !age %in% c("999", "998"),
    !employment %in% c("9", "8", "-1", "998"),
    !education %in% c("10", "99", "-1", "98"),
    !state %in% c("NA")
  )

# --- Fix Urban/Rural coding inconsistency in 2008 (3-point → 2-point scale) ---
survey <- survey %>%
  mutate(`Urban/Rural` = case_when(
    year == "2008" & `Urban/Rural` == "2" ~ "1",
    year == "2008" & `Urban/Rural` == "3" ~ "2",
    TRUE ~ `Urban/Rural`
  ))

# --- Harmonise education: recode "1" to "0" (no education) ---
survey <- survey %>%
  mutate(education = case_when(
    education == "1" ~ "0",
    TRUE ~ education
  ))

# --- Fix 1999 employment coding (0/1 were flipped) ---
survey <- survey %>%
  filter(!(year == "1999" & employment == "2")) %>%
  mutate(employment = ifelse(year == "1999" & employment == "0", "1",
                      ifelse(year == "1999" & employment == "1", "0", employment)))

# --- Harmonise employment: collapse categories into employed (1) / not employed (0) ---
survey <- survey %>%
  mutate(employment = case_when(
    year %in% c("2001","2003","2005","2008","2012","2015","2018","2021","2022") &
      employment %in% c("2","3","4","5") ~ "1",
    year %in% c("2001","2003","2005","2008","2012","2015","2018","2021","2022") &
      employment %in% c("0","1") ~ "0",
    TRUE ~ employment
  ))

# --- Standardise state name variants ---
survey <- survey %>%
  mutate(state = ifelse(state == "Abuja",       "FCT",
                 ifelse(state == "Akwa Ibom",   "Akwa-Ibom",
                 ifelse(state == "Cross River", "Cross-River", state)))) %>%
  filter(!state == "NA")

# --- Convert to correct types ---
survey <- survey %>%
  mutate(
    ID                    = as.numeric(gsub("[^0-9]", "", ID)),
    `support for democracy` = as.integer(`support for democracy`),
    `Urban/Rural`         = as.factor(`Urban/Rural`),
    age                   = as.integer(age),
    gender                = as.factor(gender),
    education             = as.factor(education),
    employment            = as.factor(employment),
    state                 = as.factor(state),
    year                  = as.integer(year)
  )

# --- Save cleaned survey ---
write.csv(survey, file = "~/Desktop/Dissertation /survey.csv", row.names = FALSE)
```


# ============================================================
# SECTION 4: CRIME DATA PREPARATION
# ============================================================

```{r prep-crime-data}
# --- Filter to Nigeria, 1999–2022 ---
crime <- crime %>%
  filter(country == "Nigeria", year >= 1999, year <= 2022, state != "") %>%
  select(state = admin1, year, event_type, disorder_type, fatalities)

# --- Aggregate: one row per state-year with event type counts & fatalities ---
crime_agg <- crime %>%
  group_by(state, year) %>%
  summarise(
    total_fatalities                     = sum(fatalities, na.rm = TRUE),
    total_events                         = n(),
    Battles_count                        = sum(event_type == "Battles"),
    Riots_count                          = sum(event_type == "Riots"),
    Protests_count                       = sum(event_type == "Protests"),
    Violence_against_civilians_count     = sum(event_type == "Violence against civilians"),
    Explosions_Remote_violence_count     = sum(event_type == "Explosions/Remote violence"),
    Strategic_developments_count        = sum(event_type == "Strategic developments"),
    Fatalities_Battles                   = sum(fatalities * (event_type == "Battles"),   na.rm = TRUE),
    Fatalities_Riots                     = sum(fatalities * (event_type == "Riots"),     na.rm = TRUE),
    Fatalities_Protests                  = sum(fatalities * (event_type == "Protests"),  na.rm = TRUE),
    Fatalities_Violence_against_civilians = sum(fatalities * (event_type == "Violence against civilians"), na.rm = TRUE),
    Fatalities_Explosions_Remote_violence = sum(fatalities * (event_type == "Explosions/Remote violence"), na.rm = TRUE),
    Fatalities_Strategic_developments    = sum(fatalities * (event_type == "Strategic developments"), na.rm = TRUE),
    .groups = "drop"
  )

# --- Add yearly national totals for relative comparisons ---
yearly_summary <- crime_agg %>%
  group_by(year) %>%
  summarise(
    yearly_total_fatalities = sum(total_fatalities),
    yearly_total_events     = sum(total_events),
    .groups = "drop"
  )

crime_agg <- crime_agg %>%
  left_join(yearly_summary, by = "year")

write.csv(crime_agg, file = "crime_agg.csv", row.names = FALSE)
```


# ============================================================
# SECTION 5: GEOGRAPHIC DATA & FINAL DATASET ASSEMBLY
# ============================================================

```{r build-final-dataset}
# --- Load Nigeria state shapefiles ---
nigeria_states <- ne_states(country = "Nigeria", returnclass = "sf") %>%
  mutate(name = recode(name,
    "Akwa Ibom"                  = "Akwa-Ibom",
    "Cross River"                = "Cross-River",
    "Federal Capital Territory"  = "FCT")) %>%
  rename(state = name) %>%
  select(state, longitude, latitude)

# --- Add support-for-democracy percentage per state-year ---
survey <- survey %>%
  group_by(state, year) %>%
  mutate(
    total_response       = n(),
    democracy_supporters = sum(support.for.democracy == 3),
    support_percentage   = democracy_supporters / total_response * 100
  ) %>%
  ungroup()

# --- Join geographic coordinates ---
survey <- left_join(survey, nigeria_states, by = "state")

# --- Merge with aggregated crime data ---
final_dataset <- survey %>%
  left_join(crime_agg, by = c("state", "year"))

write.csv(final_dataset, file = "final_dataset.csv", row.names = FALSE)
```


# ============================================================
# SECTION 6: EXPLORATORY VISUALISATIONS
# ============================================================

```{r plot-crime-fatalities-over-time}
# --- Fatalities by event type over time ---
crime_plot <- crime %>%
  group_by(year, event_type) %>%
  summarise(total_fatalities = sum(fatalities), .groups = "drop")

ggplot(crime_plot, aes(x = year, y = total_fatalities, colour = event_type)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = seq(1999, 2022, by = 3)) +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Fatalities by Event Type in Nigeria (1999–2022)",
       x = "Year", y = "Number of Fatalities", colour = "Event Type") +
  theme_minimal() +
  theme(legend.position = "bottom", plot.title = element_text(hjust = 0.5, face = "bold"))
```

```{r plot-parallel-trends}
# --- DiD treatment assignment ---
merged_data <- final_dataset %>%
  mutate(
    treated   = ifelse(state %in% c("Borno", "Yobe", "Adamawa", "Bauchi", "Gombe", "Taraba"), 1, 0),
    post_2014 = ifelse(year >= 2014, 1, 0),
    region    = ifelse(treated == 1, "North East (Treated)", "Other (Control)")
  )

# --- Pre-trend plot (pre-2014 only) ---
pre_trend_data <- merged_data %>%
  filter(year < 2014) %>%
  group_by(region, year) %>%
  summarise(avg_support = mean(support.for.democracy, na.rm = TRUE), .groups = "drop")

ggplot(pre_trend_data, aes(x = year, y = avg_support, color = region)) +
  geom_line(linewidth = 1) + geom_point() +
  scale_color_manual(values = c("North East (Treated)" = "red", "Other (Control)" = "blue")) +
  labs(title = "Pre-Trend Check: Support for Democracy (Pre-2014)",
       y = "Average Support (1–3 scale)", x = "Year", color = "Region") +
  theme_minimal()

# --- Full timeline parallel trends plot (from 2005) ---
full_trends <- merged_data %>%
  group_by(region, year) %>%
  summarise(
    avg_support = mean(support.for.democracy, na.rm = TRUE),
    se          = sd(support.for.democracy, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

full_trends %>%
  filter(year >= 2005) %>%
  ggplot(aes(x = year, y = avg_support, color = region)) +
  geom_line(linewidth = 1) + geom_point() +
  geom_errorbar(aes(ymin = avg_support - 1.96*se, ymax = avg_support + 1.96*se),
                width = 0.2, alpha = 0.5) +
  geom_vline(xintercept = 2014, linetype = "dashed", color = "red", linewidth = 1) +
  annotate("text", x = 2015, y = max(full_trends$avg_support),
           label = "Conflict Escalation (2014)", hjust = 0, color = "red") +
  scale_color_manual(values = c("North East (Treated)" = "#E41A1C", "Other (Control)" = "#377EB8")) +
  labs(title = "Support for Democracy: North East vs. Other Regions (2005–2022)",
       y = "Average Support (1–3 Scale)", x = "Year", color = "Region") +
  theme_minimal() + theme(legend.position = "bottom")
```


# ============================================================
# SECTION 7: MAIN ECONOMETRIC ANALYSIS — DIFFERENCE-IN-DIFFERENCES
# ============================================================

```{r did-sample}
# --- Restrict to 2005 onwards (data quality) ---
filtered_data <- merged_data %>%
  filter(year >= 2005) %>%
  mutate(year = as.factor(year))
```

```{r main-did-model}
# --- Two-way fixed effects DiD (state + year FE, clustered SEs at state level) ---
did_model <- feols(
  support.for.democracy ~ treated * post_2014 | state + year,
  data    = filtered_data,
  cluster = ~state
)
summary(did_model)
```

```{r event-study}
# --- Event study: year-by-year treatment effects relative to 2012 ---
event_model <- feols(
  support.for.democracy ~ i(year, treated, ref = "2012") | state + year,
  data    = filtered_data,
  cluster = ~state
)

iplot(event_model,
      main = "Event Study: Conflict Exposure and Support for Democracy",
      xlab = "Year",
      ylab = "Effect on Support for Democracy (relative to 2012)",
      col  = "darkred")
abline(h = 0, lty = 2)
```


# ============================================================
# SECTION 8: ROBUSTNESS CHECKS
# ============================================================

```{r robustness-restore-numeric-year}
# Restore numeric year for robustness manipulations
merged_data <- merged_data %>%
  mutate(year = as.numeric(as.character(year)))
```

```{r placebo-tests}
# --- Placebo 1: treat 2008 as the post period ---
placebo_2008 <- feols(
  support.for.democracy ~ treated * I(year >= 2008) | state + year,
  data = merged_data, cluster = ~state
)
summary(placebo_2008)

# --- Placebo 2: treat 2012 as the post period ---
placebo_2012 <- feols(
  support.for.democracy ~ treated * I(year >= 2012) | state + year,
  data = merged_data, cluster = ~state
)
summary(placebo_2012)
```

```{r alternative-clustering}
# --- Alternative clustering: geopolitical zone level ---
merged_data <- merged_data %>%
  mutate(zone = case_when(
    state %in% c("Borno","Yobe","Adamawa","Bauchi","Gombe","Taraba")          ~ "North East",
    state %in% c("Jigawa","Kaduna","Kano","Katsina","Kebbi","Sokoto","Zamfara") ~ "North West",
    state %in% c("Benue","Kogi","Kwara","Nasarawa","Niger","Plateau","FCT")   ~ "North Central",
    state %in% c("Abia","Anambra","Ebonyi","Enugu","Imo")                     ~ "South East",
    state %in% c("Akwa-Ibom","Bayelsa","Cross-River","Delta","Edo","Rivers")  ~ "South South",
    state %in% c("Ekiti","Lagos","Ogun","Ondo","Osun","Oyo")                  ~ "South West"
  ))

did_cluster_zone <- feols(
  support.for.democracy ~ treated * post_2014 | state + year,
  data = merged_data, cluster = ~zone
)
summary(did_cluster_zone)
```

```{r restricted-sample}
# --- Drop borderline/ambiguous states (Bauchi & Taraba) ---
did_restricted <- feols(
  support.for.democracy ~ treated * post_2014 | state + year,
  data    = merged_data %>% filter(!state %in% c("Bauchi", "Taraba")),
  cluster = ~state
)
summary(did_restricted)
```

```{r intensity-based-treatment}
# --- Continuous treatment: fatalities per 1,000 interacted with post-2014 ---
merged_data <- merged_data %>%
  mutate(fatalities_1000 = yearly_total_fatalities / 1000)

did_intensity <- feols(
  support.for.democracy ~ fatalities_1000 * post_2014,
  data = merged_data
)
summary(did_intensity)
```


# ============================================================
# SECTION 9: HETEROGENEITY ANALYSIS
# ============================================================

```{r heterogeneity-models}
# --- Education (secondary or above = 1) ---
edu_model <- feols(
  support.for.democracy ~ treated * post_2014 * I(as.numeric(as.character(education)) >= 5) | state + year,
  cluster = ~state, data = merged_data
)

# --- Gender ---
gender_model <- feols(
  support.for.democracy ~ treated * post_2014 * gender | state + year,
  cluster = ~state, data = merged_data
)

# --- Urban vs Rural ---
urban_model <- feols(
  support.for.democracy ~ treated * post_2014 * Urban.Rural | state + year,
  cluster = ~state, data = merged_data
)

# --- Employment ---
employment_model <- feols(
  support.for.democracy ~ treated * post_2014 * employment | state + year,
  cluster = ~state, data = merged_data
)

# --- Age groups ---
merged_data <- merged_data %>%
  mutate(age_group = case_when(
    age < 30           ~ "Youth",
    age >= 30 & age < 50 ~ "Middle",
    TRUE               ~ "Older"
  ))

age_model <- feols(
  support.for.democracy ~ treated * post_2014 * age_group | state + year,
  cluster = ~state, data = merged_data
)

# --- Combined heterogeneity table ---
modelsummary(
  list("Education" = edu_model, "Gender" = gender_model,
       "Urban/Rural" = urban_model, "Employment" = employment_model,
       "Age Groups" = age_model),
  stars     = TRUE,
  gof_omit  = "AIC|BIC|Log.Lik",
  title     = "Table: Heterogeneity of Conflict Exposure Effects on Democratic Support"
)
```


# ============================================================
# SECTION 10: SUMMARY STATISTICS TABLE
# ============================================================

```{r summary-statistics}
# --- Create 4-group variable: Treated/Control x Pre/Post-2014 ---
filtered_data_num <- merged_data %>%
  filter(year >= 2005) %>%
  mutate(group = case_when(
    treated == 1 & post_2014 == 0 ~ "Treated_Pre2014",
    treated == 1 & post_2014 == 1 ~ "Treated_Post2014",
    treated == 0 & post_2014 == 0 ~ "Control_Pre2014",
    treated == 0 & post_2014 == 1 ~ "Control_Post2014"
  ))

# --- Group-level summary stats ---
group_stats <- filtered_data_num %>%
  group_by(group) %>%
  summarise(
    Mean_Age            = mean(age, na.rm = TRUE),
    SD_Age              = sd(age, na.rm = TRUE),
    Percent_Female      = mean(gender == 1, na.rm = TRUE) * 100,
    Percent_Unemployed  = mean(employment == 0, na.rm = TRUE) * 100,
    Percent_Urban       = mean(Urban.Rural == 1, na.rm = TRUE) * 100,
    Percent_Pro_Demo    = mean(support.for.democracy == 3, na.rm = TRUE) * 100,
    Percent_Educated    = mean(as.numeric(as.character(education)) >= 5, na.rm = TRUE) * 100,
    Mean_Fatalities     = mean(yearly_total_fatalities, na.rm = TRUE),
    SD_Fatalities       = sd(yearly_total_fatalities, na.rm = TRUE),
    Mean_Events         = mean(yearly_total_events, na.rm = TRUE),
    SD_Events           = sd(yearly_total_events, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(factor(group, levels = c("Treated_Pre2014","Treated_Post2014",
                                    "Control_Pre2014","Control_Post2014")))

# --- p-values: treated vs control, pre vs post ---
pvals_conflict <- tibble(
  Age        = t.test(age ~ treated,                              data = filtered_data_num)$p.value,
  Female     = t.test(gender == 1 ~ treated,                     data = filtered_data_num)$p.value,
  Unemp      = t.test(employment == 0 ~ treated,                 data = filtered_data_num)$p.value,
  Urban      = t.test(Urban.Rural == 1 ~ treated,                data = filtered_data_num)$p.value,
  Democracy  = t.test(support.for.democracy == 3 ~ treated,      data = filtered_data_num)$p.value,
  Educated   = t.test(as.numeric(as.character(education)) >= 5 ~ treated, data = filtered_data_num)$p.value,
  Fatalities = t.test(yearly_total_fatalities ~ treated,         data = filtered_data_num)$p.value,
  Events     = t.test(yearly_total_events ~ treated,             data = filtered_data_num)$p.value
)

pvals_post <- tibble(
  Age        = t.test(age ~ post_2014,                              data = filtered_data_num)$p.value,
  Female     = t.test(gender == 1 ~ post_2014,                     data = filtered_data_num)$p.value,
  Unemp      = t.test(employment == 0 ~ post_2014,                 data = filtered_data_num)$p.value,
  Urban      = t.test(Urban.Rural == 1 ~ post_2014,                data = filtered_data_num)$p.value,
  Democracy  = t.test(support.for.democracy == 3 ~ post_2014,      data = filtered_data_num)$p.value,
  Educated   = t.test(as.numeric(as.character(education)) >= 5 ~ post_2014, data = filtered_data_num)$p.value,
  Fatalities = t.test(yearly_total_fatalities ~ post_2014,         data = filtered_data_num)$p.value,
  Events     = t.test(yearly_total_events ~ post_2014,             data = filtered_data_num)$p.value
)

# --- Final formatted table ---
summary_table <- tibble(
  Characteristic = c("Mean Age (SD)", "% Female", "% Unemployed", "% Urban",
                     "% Support Democracy", "% Educated (secondary+)",
                     "Fatalities (Mean, SD)", "Total Events (Mean, SD)"),
  `Treated (Pre-2014)`  = sprintf_stats(group_stats, 1),
  `Treated (Post-2014)` = sprintf_stats(group_stats, 2),
  `Control (Pre-2014)`  = sprintf_stats(group_stats, 3),
  `Control (Post-2014)` = sprintf_stats(group_stats, 4),
  `p (Conflict)`        = format_pvals(pvals_conflict),
  `p (Post-2014)`       = format_pvals(pvals_post)
)

kable(summary_table, caption = "Table 1: Sample Characteristics by Conflict Exposure and Time Period")
```
